---
title: "Running a Study: From Setup to Batch Results"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running a Study: From Setup to Batch Results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

`llmtextcoder` is an R package for **LLM-assisted qualitative text coding**.
In psychology and social science research, participants often provide
free-text responses — descriptions of experiences, thoughts, or feelings —
that need to be rated along theoretically motivated dimensions. Traditionally
this is done by trained human raters: time-consuming, expensive, and subject
to fatigue and inconsistency.

`llmtextcoder` automates this process. You write a scoring rubric as a plain
text template, supply your data, and the package sends each response to an
LLM (currently OpenAI's API), receiving structured scores in return. The
scores are saved to a flat CSV ready for statistical analysis.

All examples in this vignette use `varieties_testimonials()`, a bundled
dataset of 47 first-person psychological narratives drawn from William James's
*The Varieties of Religious Experience* (1902, public domain). The texts are
real, psychologically varied, and ready to run — you can copy every code block
and execute it immediately after supplying an API key.

### A healthy workflow

A well-run study follows these stages, in order:

1. **Write a rubric** and verify it compiles correctly (free — no API calls).
2. **Screen and redact** participant data for identifying information before
   anything leaves your computer.
3. **Pilot** on a small batch to confirm the rubric is being interpreted as
   intended.
4. **Score** the full dataset, with automatic skip logic if the run is
   interrupted.

The sections below walk through each stage.

---

## 1. Installation

```{r}
# install.packages("remotes")
remotes::install_github("eladzlot/llmtextcoder")
library(llmtextcoder)
```

## 2. Store your API key

You need an OpenAI API key. Sign in at <https://platform.openai.com>, go to
*API keys*, and create one.

**Important — account tier and data privacy.**
Not all OpenAI accounts offer the same privacy protections. On consumer
accounts (including the free tier and standard pay-as-you-go), OpenAI may
use submitted content to improve its models. Only enterprise-tier accounts
and institutional agreements (such as ChatGPT Edu) contractually prohibit
this. If you are working with participant data, confirm with your lab or
institution which account type you are authorized to use before proceeding.

Never paste your key into a script. Instead, create a file called `.env`
somewhere safe (typically your project root) and add:

```
OPENAI_API_KEY=sk-proj-...
```

Add `.env` to your `.gitignore`. Load it at the top of every session:

```{r}
readRenviron(".env")
nzchar(Sys.getenv("OPENAI_API_KEY"))  # should be TRUE
```

## 3. Write your rubric

A rubric is a string with `{{placeholder}}` markers where participant data
gets injected. Two requirements:

1. Each placeholder name must match a column in your data frame.
2. The rubric must instruct the model to respond in JSON (OpenAI requirement).

You can define a rubric inline or load it from a `.txt` file.

**Inline** (convenient for interactive work and vignettes):

```{r}
rubric <- "
You are an expert clinical psychologist coding first-person narrative accounts.
Rate the excerpt below on three psychological dimensions.

Dimension 1 — EMOTIONAL VALENCE (1-5)
1 = strongly positive affect  5 = strongly negative affect

Dimension 2 — HOPELESSNESS (1-5)
1 = clear hope or agency  5 = complete hopelessness, no future orientation

Dimension 3 — SOMATIC DISTRESS (1-5)
1 = no bodily complaints  5 = pervasive physical suffering described

Respond ONLY with a JSON object, nothing else:
{\"valence\": <1-5>, \"hopelessness\": <1-5>, \"somatic_distress\": <1-5>}

Excerpt:
{{text}}
"
```

**From a file** (recommended once you start collecting data):

```{r}
rubric <- paste(readLines("prompts/clinical_v1.txt"), collapse = "\n")
```

**The rubric is your scientific instrument.** Once you have collected data
against a version, never edit it in place. Save revisions under a new name
(`clinical_v2.txt`) so results always trace back to the exact rubric that
produced them.

## 4. Load the example data and inspect the prompt

```{r}
df <- varieties_testimonials()
head(df[, c("id", "theme", "word_count")])
#>        id        theme word_count
#> 1 james_01     presence        223
#> 2 james_02  faith_doubt        473
#> 3 james_03     mystical        229
#> 4 james_04     mystical        248
#> 5 james_05     mystical        541
#> 6 james_06 illness_healing       393
```

Before spending any API credits, verify that the prompt assembles correctly
for one row:

```{r}
entry <- df[df$id == "james_10", ]   # depression theme
preview_prompt(rubric, list(text = entry$text))
```

This prints the full compiled prompt with character and word counts. Read it
as the model will see it — confirm the text landed in the right place and the
rubric instructions are complete.

## 5. Score a single text interactively

Once the prompt looks right, test the full round-trip with one text:

```{r}
params <- run_params(model = "gpt-4o", temperature = 0)

result <- score_one(rubric, list(text = entry$text), params)
print_result(result$raw, result$scores)
#> Raw JSON:
#> {"valence": 5, "hopelessness": 5, "somatic_distress": 4}
#>
#> Parsed scores:
#>   valence         : 5
#>   hopelessness    : 5
#>   somatic_distress: 4
```

`print_result()` shows both the raw JSON the model returned and the parsed
scores side by side. Ask yourself:

- Are all expected dimensions present?
- Are the values in the right range?
- Does the rating make intuitive sense for this text?

Repeat with two or three contrasting texts before moving on:

```{r}
# james_33 is rated "mystical" — should score very differently
mystical <- df[df$id == "james_33", ]
result2  <- score_one(rubric, list(text = mystical$text), params)
print_result(result2$raw, result2$scores)
```

This sanity check costs a few API calls but can save you from discovering a
rubric problem after scoring 500 participants.

## 6. Prepare your data

`score_many()` expects a data frame with:

| Column | Description |
|--------|-------------|
| `id`   | A unique identifier for each row |
| *(placeholder columns)* | One column per `{{placeholder}}` in your rubric |

`varieties_testimonials()` already has the right shape — an `id` column and a
`text` column matching the `{{text}}` placeholder in the rubric above:

```{r}
# Use the full dataset or a subset
batch <- df[, c("id", "text")]
nrow(batch)   # 47
```

For your own data:

```{r}
batch <- read.csv("raw/participants.csv", stringsAsFactors = FALSE)
head(batch[, c("id", "text")])
```

The `id` column is what the skip logic uses to avoid rescoring rows that have
already been processed.

## 7. Scan and redact identifying information

Before sending any text to an external API, scan for potentially identifying
information (PII). The James testimonials are public domain and contain no
PII, so we skip the check for them:

```{r}
score_many(batch, rubric, "clinical_v1", params,
           pii_check = FALSE, output_dir = "results")
```

For your own participant data, always run the scan:

```{r}
pii_result <- scan_pii(batch)
```

`scan_pii()` separates findings into two tiers and prints a ready-to-use
action plan:

```
PII scan: 4 match(es) across 3 text(s)

Auto-redactable — use redact_pii() to clear:
  [P089] email
    "contact me at user@example.com"

Needs review — use redact_words() per row:
  [P023] name_disclosure #1
    "My name is Sarah Johnson and I…"
```

### Step 1 — Auto-redact structural PII

**Never edit your source data file.** Raw data is a scientific record; all
changes must live in code. `redact_pii()` replaces structural PII (emails,
phones, Prolific IDs, etc.) globally:

```{r}
batch <- redact_pii(batch)
```

### Step 2 — Redact disclosure phrases per row

Disclosure patterns ("my name is", "I live in") capture only the trigger
phrase, not the identifying word that follows. How many words to remove is a
judgment call per row.

```{r}
result <- scan_pii(batch)
pii_code(result)          # prints suggested redact_words() calls
```

Paste the output into your script and fill in `n_words`:

```{r}
batch <- redact_words(batch, id = "P023", pattern = "name_disclosure", n_words = 2)
batch <- redact_words(batch, id = "P041", pattern = "location_phrase",  n_words = 3)
```

## 8. Choose an output directory

`score_many()` and `status()` both take an `output_dir` argument. Results are
written to `<output_dir>/<output_name>.csv` — so the filename always encodes
which rubric version produced it:

```{r}
output_dir <- "results"   # or "data", "output", an absolute path, etc.
```

The directory is created automatically if it does not exist.

## 9. Pilot run: score a small batch first

Always run a pilot before committing the full dataset. Use `n` to limit how
many rows are processed:

```{r}
score_many(batch, rubric, "clinical_v1", params,
           n = 5, pii_check = FALSE, output_dir = output_dir)
```

Open `results/clinical_v1.csv` and check:

- Are all expected score columns present (`valence`, `hopelessness`,
  `somatic_distress`)?
- Do the scores look reasonable across varied texts?
- Is there any evidence the rubric is being misread?

If something is wrong, fix the rubric and change `output_name` to
`"clinical_v2"` before re-piloting. Never mix rubric versions in one CSV.

## 10. Full batch run

Once the pilot looks good, run the full dataset. Rows already scored in the
pilot are automatically skipped:

```{r}
score_many(batch, rubric, "clinical_v1", params,
           pii_check = FALSE, output_dir = output_dir)
```

A progress bar shows completion in real time. Individual failures are caught
per-row and written to `results/clinical_v1_errors.csv` — the run continues
rather than stopping. A summary is printed at the end:

```
Scoring 42 row(s) → results/clinical_v1.csv
[==========================================] 42/42 (100%)
Done. 41 scored, 1 failed — see results/clinical_v1_errors.csv.
```

Check status at any point:

```{r}
status(batch, "clinical_v1", params, output_dir = output_dir)
# Rubric     : clinical_v1
# Model      : gpt-4o
# Output dir : results
# Total      : 47
# Scored     : 46
# Failed     :  1
# Pending    :  0
```

## 11. Handling interruptions

If the run is interrupted (network error, crash, power loss), just run the
same command again. Completed rows are skipped, failed rows are retried:

```{r}
# Safe to run again at any time
score_many(batch, rubric, "clinical_v1", params,
           pii_check = FALSE, output_dir = output_dir)
```

To process only the next N rows:

```{r}
score_many(batch, rubric, "clinical_v1", params,
           n = 10, pii_check = FALSE, output_dir = output_dir)
```

## 12. Large datasets: async batch scoring

For datasets of several hundred texts or more, the
[OpenAI Batch API](https://platform.openai.com/docs/guides/batch) is the
recommended approach. It submits all requests as a single asynchronous job,
returns results within 24 hours, and costs **50% less** than the synchronous
API. `score_many()` calls the API once per text in sequence — fine for pilots
and moderate datasets, but slow and expensive at scale.

The batch API still takes a template **file path** (the path is used to name
the state file on disk):

### Submit the job

```{r}
submit_batch(batch, "prompts/clinical_v1.txt", params, output_dir = output_dir)
```

This builds a JSONL file of all pending rows (same skip logic as
`score_many()`), uploads it to OpenAI, and submits the batch job. A state
file `<output_dir>/clinical_v1_batch_<model>.json` is written immediately.

### Collect results

When you return (anywhere from a few minutes to 24 hours later), call:

```{r}
collect_batch("prompts/clinical_v1.txt", params, output_dir = output_dir)
```

If the job is still running, the current progress is printed and the function
returns `NULL` — call again later. When complete, results are parsed and
written to the same CSV as `score_many()`.

### Choosing between score_many() and submit_batch()

| | `score_many()` | `submit_batch()` |
|---|---|---|
| Results available | Immediately, row by row | Within 24 hours |
| Cost | Standard rate | 50% discount |
| Good for | Pilots, small datasets | Large datasets (200+ texts) |
| Interruption recovery | Resume with same call | Call `collect_batch()` later |

## 13. Reading and using your results

Each row of `results/clinical_v1.csv` contains:

| Column | Description |
|--------|-------------|
| `id` | The identifier from your input data |
| `valence`, `hopelessness`, `somatic_distress` | One column per JSON key returned by the model |
| `raw` | The exact JSON string from the model (for auditing) |
| `prompt_version` | The `output_name` used in `score_many()` |
| `model` | The model name (e.g. `gpt-4o`) |
| `temperature` | The temperature setting used |
| `scored_at` | ISO 8601 timestamp |

Input texts are not stored in the output CSV — only `id` links back to your
source data. This prevents accidental leakage when sharing results, and keeps
the schema flat even when your rubric uses multiple input columns.

```{r}
results <- read.csv("results/clinical_v1.csv", stringsAsFactors = FALSE)

# Aggregate scores
results$severity <- rowMeans(results[, c("valence", "hopelessness",
                                         "somatic_distress")])

# Compare themes from the original dataset
merged <- merge(results[, c("id", "severity")],
                df[,      c("id", "theme")],
                by = "id")
tapply(merged$severity, merged$theme, mean)
```

The provenance columns (`prompt_version`, `model`, `scored_at`) make every
row self-describing — a CSV can be archived or shared and its origin is
always recoverable.
