---
title: "Running a Study: From Setup to Batch Results"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running a Study: From Setup to Batch Results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

`llmtextcoder` is an R package for **LLM-assisted qualitative text coding**.
In psychology and social science research, participants often provide
free-text responses — descriptions of experiences, thoughts, or feelings —
that need to be rated along theoretically motivated dimensions. Traditionally
this is done by trained human raters: time-consuming, expensive, and subject
to fatigue and inconsistency.

`llmtextcoder` automates this process. You write a scoring rubric as a plain
text template, supply your participant data, and the package sends each
response to an LLM (currently OpenAI's API), receiving structured scores in
return. The scores are saved to a flat CSV ready for statistical analysis.

### A healthy workflow

A well-run study follows these stages, in order:

1. **Write a rubric** and verify it compiles correctly (free — no API calls).
2. **Screen and redact** participant data for identifying information before
   anything leaves your computer.
3. **Pilot** on a small batch to confirm the rubric is being interpreted as
   intended.
4. **Score** the full dataset, with automatic skip logic if the run is
   interrupted.

The sections below walk through each stage.

---

## 1. Installation

```{r}
# install.packages("remotes")
remotes::install_github("eladzlot/llmtextcoder")
library(llmtextcoder)
```

## 2. Store your API key

You need an OpenAI API key. Sign in at <https://platform.openai.com>, go to
*API keys*, and create one.

**Important — account tier and data privacy.**
Not all OpenAI accounts offer the same privacy protections. On consumer
accounts (including the free tier and standard pay-as-you-go), OpenAI may
use submitted content to improve its models. Only enterprise-tier accounts
and institutional agreements (such as ChatGPT Edu) contractually prohibit
this. If you are working with participant data, confirm with your lab or
institution which account type you are authorized to use before proceeding.

Never paste your key into a script. Instead, create a file called `.env`
somewhere safe (typically your project root) and add:

```
OPENAI_API_KEY=sk-proj-...
```

Add `.env` to your `.gitignore`. Load it at the top of every session:

```{r}
readRenviron(".env")
nzchar(Sys.getenv("OPENAI_API_KEY"))  # should be TRUE
```

## 3. Write your rubric template

A rubric is a plain `.txt` file. The only requirements are:

1. It contains `{{text}}` — the placeholder where participant text gets injected.
2. It instructs the model to respond in JSON (required for structured output).

Example — save this as `rubric_v1.txt`:

```
You are an expert rater. Score the following text on two dimensions,
each on a 1–5 scale.

Dimension 1 — Clarity (1 = very unclear, 5 = very clear)
Dimension 2 — Emotional tone (1 = very negative, 5 = very positive)

Respond ONLY with a JSON object in this exact format and nothing else:
{"clarity": <1-5>, "emotional_tone": <1-5>}

Text to rate:
{{text}}
```

**The rubric is your scientific instrument.** Once you have collected data
against a version, never edit it in place. Save revisions under a new name
(`rubric_v2.txt`) so your output files always trace back to the exact rubric
that produced them.

## 4. Inspect the compiled prompt (no API cost)

Before spending any API credits, verify that the prompt assembles correctly:

```{r}
template_path    <- "rubric_v1.txt"
participant_text <- "The instructions were easy to follow and I felt great."

preview_prompt(read_template(template_path), participant_text)
```

This prints the full prompt with character and word counts. Read it as the
model will see it — confirm the participant text landed in the right place and
the rubric instructions are complete.

## 5. Score a single text interactively

Once the prompt looks right, test the full round-trip with one text:

```{r}
params <- run_params(model = "gpt-4o", temperature = 0)

result <- score_one(template_path, participant_text, params)
print_result(result$raw, result$scores)
```

`print_result()` shows both the raw JSON the model returned and the parsed
scores side by side. Ask yourself:

- Are all expected dimensions present?
- Are the values in the right range?
- Does the rating make intuitive sense for this text?

Repeat with two or three contrasting texts before moving on. This sanity check
costs a few API calls but can save you from discovering a rubric problem after
scoring 500 participants.

## 6. Prepare your data

`score_many()` expects a data frame with at minimum two columns:

| Column | Description |
|--------|-------------|
| `id`   | A unique identifier for each row (e.g. a participant ID) |
| `text` | The text to score |

```{r}
df <- read.csv("participants.csv", stringsAsFactors = FALSE)
head(df[, c("id", "text")])
```

The `id` column is what the skip logic uses to avoid rescoring rows that have
already been processed.

## 7. Scan and redact identifying information

Before sending any text to an external API, scan for potentially identifying
information (PII). This is especially important when participants provided
free-text responses.

```{r}
pii_result <- scan_pii(df)
```

`scan_pii()` separates findings into two tiers and prints a ready-to-use
action plan:

```
PII scan: 4 match(es) across 3 text(s)

Auto-redactable — use redact_pii() to clear:
  [P089] email
    "contact me at user@example.com"

Needs review — use redact_words() per row:
  [P023] name_disclosure #1
    "My name is Sarah Johnson and I…"
  [P023] name_disclosure #2
    "my friend's name is David and…"
  [P041] location_phrase #1
    "I live in Tel Aviv near the old…"

Suggested redact_words() calls — fill in n_words:
  df <- redact_words(df, id = "P023", pattern = "name_disclosure", n_words = ?, occurrence = 1)
  df <- redact_words(df, id = "P023", pattern = "name_disclosure", n_words = ?, occurrence = 2)
  df <- redact_words(df, id = "P041", pattern = "location_phrase", n_words = ?)
```

### Step 1 — Auto-redact structural PII

**Never edit your source data file.** Raw data is a scientific record; all
changes must live in code. `redact_pii()` replaces structural PII (emails,
phones, Prolific IDs, etc.) globally using the detection regex — the original
values never appear in your script:

```{r}
df <- redact_pii(df)
```

### Step 2 — Redact disclosure phrases per row

Disclosure patterns ("my name is", "I live in") capture only the trigger
phrase, not the name or place that follows. How many words to remove is a
judgment call per row.

Use `pii_code()` to get all suggested calls as a string you can paste
directly into your script — or copy to the clipboard in one step:

```{r}
result <- scan_pii(df)
pii_code(result)          # prints the suggested calls
# clipr::write_clip(pii_code(result))  # copies them to the clipboard
```

Paste the output into your script and fill in `n_words` for each row:

```{r}
df <- redact_words(df, id = "P023", pattern = "name_disclosure", n_words = 2, occurrence = 1)
df <- redact_words(df, id = "P023", pattern = "name_disclosure", n_words = 1, occurrence = 2)
df <- redact_words(df, id = "P041", pattern = "location_phrase",  n_words = 3)
```

`redact_words()` keeps the trigger phrase ("My name is") and replaces only
the words that follow it with `[name disclosure redacted]`. When `occurrence`
is omitted, all matches in that row are replaced with the same word count.

### Skipping the check on subsequent runs

`score_many()` runs `scan_pii()` automatically (default `pii_check = TRUE`).
Once redaction is complete, set `pii_check = FALSE`:

```{r}
score_many(df, template_path, params,
           pii_check = FALSE, output_dir = output_dir)
```

### Customising the patterns

Add domain-specific patterns or remove noisy ones by modifying the two
pattern sets independently:

```{r}
auto <- auto_pii_patterns()
auto["national_id"] <- "\\b\\d{9}\\b"  # add Israeli national ID
auto <- auto[names(auto) != "url"]      # URLs not sensitive in this study

scan_pii(df, auto_patterns = auto)
redact_pii(df, patterns = auto)
```

## 8. Choose an output directory

`score_many()` and `status()` both take an `output_dir` argument. Results are
written to `<output_dir>/<rubric-stem>.csv` — so the filename always encodes
which rubric version produced it. Choose a directory that makes sense for your
project:

```{r}
output_dir <- "results"   # or "data", "output", an absolute path, etc.
```

The directory is created automatically if it does not exist.

## 9. Pilot run: score a small batch first

Always run a pilot before committing the full dataset. Use `n` to limit how
many rows are processed:

```{r}
score_many(df, template_path, params,
           n = 10, pii_check = FALSE, output_dir = output_dir)
```

Open `results/rubric_v1.csv` and check:

- Are all expected score columns present?
- Do the scores look reasonable across varied texts?
- Is there any evidence the rubric is being misread?

If something is wrong, fix the rubric (save it as `rubric_v2.txt`) and pilot
again before touching the full dataset.

## 10. Full batch run

Once the pilot looks good, run the full dataset. Rows already scored in the
pilot are automatically skipped:

```{r}
score_many(df, template_path, params,
           pii_check = FALSE, output_dir = output_dir)
```

A progress bar shows completion in real time. Individual failures are caught
per-row and written to `results/rubric_v1_errors.csv` — the run continues
rather than stopping. A summary is printed at the end:

```
Scoring 490 text(s) → results/rubric_v1.csv
[==========================================] 490/490 (100%)
Done. 488 scored, 2 failed — see results/rubric_v1_errors.csv.
```

Check status at any point:

```{r}
status(df, template_path, params, output_dir = output_dir)
# Rubric     : rubric_v1.txt
# Model      : gpt-4o
# Output dir : results
# Total      : 500
# Scored     : 488
# Failed     :   2
# Pending    :  10
```

## 11. Handling interruptions

If the run is interrupted (network error, crash, power loss), just run the
same command again. Completed rows are skipped, failed rows are retried:

```{r}
# Safe to run again at any time
score_many(df, template_path, params, output_dir = output_dir)
```

To process only the next batch of N rows:

```{r}
score_many(df, template_path, params, n = 50, output_dir = output_dir)
```

## 12. Large datasets: async batch scoring

For datasets of several hundred texts or more, the
[OpenAI Batch API](https://platform.openai.com/docs/guides/batch) is the
recommended approach. It submits all requests as a single asynchronous job,
returns results within 24 hours, and costs **50% less** than the synchronous
API. `score_many()` calls the API once per text in sequence — fine for pilots
and moderate datasets, but slow and expensive at scale.

### Submit the job

```{r}
submit_batch(df, template_path, params, output_dir = output_dir)
```

This builds a JSONL file of all pending rows (same skip logic as
`score_many()`), uploads it to OpenAI, and submits the batch job. A state
file `<output_dir>/<stem>_batch_<model>.json` is written immediately — it
records the batch ID and the original texts so you can close R and come back
later without losing anything.

### Collect results

When you return (anywhere from a few minutes to 24 hours later), call:

```{r}
collect_batch(template_path, params, output_dir = output_dir)
```

If the job is still running, the current progress is printed and the function
returns `NULL` — call again later. When complete, results are parsed and
written to the same CSV as `score_many()`, failures go to `_errors.csv`, and
the state file is removed.

### Running multiple batches simultaneously

Each rubric–model combination has its own state file, so you can submit
batches for different rubrics or models at the same time:

```{r}
submit_batch(df, "rubric_v1.txt", run_params("gpt-4o"),      output_dir = output_dir)
submit_batch(df, "rubric_v1.txt", run_params("gpt-4o-mini"), output_dir = output_dir)
submit_batch(df, "rubric_v2.txt", run_params("gpt-4o"),      output_dir = output_dir)

# Later — collect each independently
collect_batch("rubric_v1.txt", run_params("gpt-4o"),      output_dir = output_dir)
collect_batch("rubric_v1.txt", run_params("gpt-4o-mini"), output_dir = output_dir)
collect_batch("rubric_v2.txt", run_params("gpt-4o"),      output_dir = output_dir)
```

### Choosing between score_many() and submit_batch()

| | `score_many()` | `submit_batch()` |
|---|---|---|
| Results available | Immediately, row by row | Within 24 hours |
| Cost | Standard rate | 50% discount |
| Good for | Pilots, small datasets | Large datasets (200+ texts) |
| Interruption recovery | Resume with same call | Call `collect_batch()` later |

## 13. Reading and using your results

Each row of the output CSV contains:

| Column | Description |
|--------|-------------|
| `id` | The participant identifier from your input data |
| `text` | The original text (useful for spot-checking) |
| *(score columns)* | One column per JSON key the model returned |
| `raw` | The exact JSON string from the model (for auditing) |
| `prompt_version` | Rubric filename stem (e.g. `rubric_v1`) |
| `model` | The model name (e.g. `gpt-4o`) |
| `temperature` | The temperature setting used |
| `scored_at` | ISO 8601 timestamp |

```{r}
results <- read.csv("results/rubric_v1.csv", stringsAsFactors = FALSE)

# Downstream analysis: aggregate scores however your study requires
results$overall <- rowMeans(results[, c("clarity", "emotional_tone")])
hist(results$overall)
```

The provenance columns (`prompt_version`, `model`, `scored_at`) make every row
self-describing — a CSV can be archived or shared and its origin is always
recoverable.
