---
title: "Running a Study: From Setup to Batch Results"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running a Study: From Setup to Batch Results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

This guide walks a new user through a complete study: from installation and
writing a rubric, to inspecting results and scoring a full dataset.

## 1. Installation

```{r}
# install.packages("remotes")
remotes::install_github("your-username/prompter")
library(llmtextcoder)
```

## 2. Store your API key

You need an OpenAI API key. Sign in at <https://platform.openai.com>, go to
*API keys*, and create one.

Never paste your key into a script. Instead, create a file called `.env`
somewhere safe (typically your project root) and add:

```
OPENAI_API_KEY=sk-proj-...
```

Add `.env` to your `.gitignore`. Load it at the top of every session:

```{r}
readRenviron(".env")
nzchar(Sys.getenv("OPENAI_API_KEY"))  # should be TRUE
```

## 3. Write your rubric template

A rubric is a plain `.txt` file. The only requirements are:

1. It contains `{{text}}` — the placeholder where participant text gets injected.
2. It instructs the model to respond in JSON (required for structured output).

Example — save this as `rubric_v1.txt`:

```
You are an expert rater. Score the following text on two dimensions,
each on a 1–5 scale.

Dimension 1 — Clarity (1 = very unclear, 5 = very clear)
Dimension 2 — Emotional tone (1 = very negative, 5 = very positive)

Respond ONLY with a JSON object in this exact format and nothing else:
{"clarity": <1-5>, "emotional_tone": <1-5>}

Text to rate:
{{text}}
```

**The rubric is your scientific instrument.** Once you have collected data
against a version, never edit it in place. Save revisions under a new name
(`rubric_v2.txt`) so your output files always trace back to the exact rubric
that produced them.

## 4. Inspect the compiled prompt (no API cost)

Before spending any API credits, verify that the prompt assembles correctly:

```{r}
template_path    <- "rubric_v1.txt"
participant_text <- "The instructions were easy to follow and I felt great."

preview_prompt(read_template(template_path), participant_text)
```

This prints the full prompt with character and word counts. Read it as the
model will see it — confirm the participant text landed in the right place and
the rubric instructions are complete.

## 5. Score a single text interactively

Once the prompt looks right, test the full round-trip with one text:

```{r}
params <- run_params(model = "gpt-4o", temperature = 0)

result <- score_one(template_path, participant_text, params)
print_result(result$raw, result$scores)
```

`print_result()` shows both the raw JSON the model returned and the parsed
scores side by side. Ask yourself:

- Are all expected dimensions present?
- Are the values in the right range?
- Does the rating make intuitive sense for this text?

Repeat with two or three contrasting texts before moving on. This sanity check
costs a few API calls but can save you from discovering a rubric problem after
scoring 500 participants.

## 6. Prepare your data

`score_many()` expects a data frame with at minimum two columns:

| Column | Description |
|--------|-------------|
| `id`   | A unique identifier for each row (e.g. a participant ID) |
| `text` | The text to score |

```{r}
df <- read.csv("participants.csv", stringsAsFactors = FALSE)
head(df[, c("id", "text")])
```

The `id` column is what the skip logic uses to avoid rescoring rows that have
already been processed.

## 7. Choose an output directory

`score_many()` and `status()` both take an `output_dir` argument. Results are
written to `<output_dir>/<rubric-stem>.csv` — so the filename always encodes
which rubric version produced it. Choose a directory that makes sense for your
project:

```{r}
output_dir <- "results"   # or "data", "output", an absolute path, etc.
```

The directory is created automatically if it does not exist.

## 8. Pilot run: score a small batch first

Always run a pilot before committing the full dataset. Use `n` to limit how
many rows are processed:

```{r}
score_many(df, template_path, params, n = 10, output_dir = output_dir)
```

Open `results/rubric_v1.csv` and check:

- Are all expected score columns present?
- Do the scores look reasonable across varied texts?
- Is there any evidence the rubric is being misread?

If something is wrong, fix the rubric (save it as `rubric_v2.txt`) and pilot
again before touching the full dataset.

## 9. Full batch run

Once the pilot looks good, run the full dataset. Rows already scored in the
pilot are automatically skipped:

```{r}
score_many(df, template_path, params, output_dir = output_dir)
```

A progress bar shows completion in real time. Individual failures are caught
per-row and written to `results/rubric_v1_errors.csv` — the run continues
rather than stopping. A summary is printed at the end:

```
Scoring 490 text(s) → results/rubric_v1.csv
[==========================================] 490/490 (100%)
Done. 488 scored, 2 failed — see results/rubric_v1_errors.csv.
```

Check status at any point:

```{r}
status(df, template_path, params, output_dir = output_dir)
# Rubric     : rubric_v1.txt
# Model      : gpt-4o
# Output dir : results
# Total      : 500
# Scored     : 488
# Failed     :   2
# Pending    :  10
```

## 10. Handling interruptions

If the run is interrupted (network error, crash, power loss), just run the
same command again. Completed rows are skipped, failed rows are retried:

```{r}
# Safe to run again at any time
score_many(df, template_path, params, output_dir = output_dir)
```

To process only the next batch of N rows:

```{r}
score_many(df, template_path, params, n = 50, output_dir = output_dir)
```

## 11. Comparing rubric versions

Save your revised rubric as `rubric_v2.txt`. Each call to `score_many()`
produces a separate file:

```{r}
score_many(df, "rubric_v1.txt", params, output_dir = output_dir)
# → results/rubric_v1.csv

score_many(df, "rubric_v2.txt", params, output_dir = output_dir)
# → results/rubric_v2.csv
```

Compare them in R:

```{r}
v1 <- read.csv("results/rubric_v1.csv", stringsAsFactors = FALSE)
v2 <- read.csv("results/rubric_v2.csv", stringsAsFactors = FALSE)

mean(v1$clarity)
mean(v2$clarity)
```

## 12. Reading and using your results

Each row of the output CSV contains:

| Column | Description |
|--------|-------------|
| `id` | The participant identifier from your input data |
| `text` | The original text (useful for spot-checking) |
| *(score columns)* | One column per JSON key the model returned |
| `raw` | The exact JSON string from the model (for auditing) |
| `prompt_version` | Rubric filename stem (e.g. `rubric_v1`) |
| `model` | The model name (e.g. `gpt-4o`) |
| `temperature` | The temperature setting used |
| `scored_at` | ISO 8601 timestamp |

```{r}
results <- read.csv("results/rubric_v1.csv", stringsAsFactors = FALSE)

# Downstream analysis: aggregate scores however your study requires
results$overall <- rowMeans(results[, c("clarity", "emotional_tone")])
hist(results$overall)
```

The provenance columns (`prompt_version`, `model`, `scored_at`) make every row
self-describing — a CSV can be archived or shared and its origin is always
recoverable.
