---
title: "Running a Study: From Setup to Batch Results"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running a Study: From Setup to Batch Results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

This guide walks a new user through a complete study: from installation and
writing a rubric, to inspecting results and scoring a full dataset.

## 1. Installation

```{r}
# install.packages("remotes")
remotes::install_github("eladzlot/llmtextcoder")
library(llmtextcoder)
```

## 2. Store your API key

You need an OpenAI API key. Sign in at <https://platform.openai.com>, go to
*API keys*, and create one.

Never paste your key into a script. Instead, create a file called `.env`
somewhere safe (typically your project root) and add:

```
OPENAI_API_KEY=sk-proj-...
```

Add `.env` to your `.gitignore`. Load it at the top of every session:

```{r}
readRenviron(".env")
nzchar(Sys.getenv("OPENAI_API_KEY"))  # should be TRUE
```

## 3. Write your rubric template

A rubric is a plain `.txt` file. The only requirements are:

1. It contains `{{text}}` — the placeholder where participant text gets injected.
2. It instructs the model to respond in JSON (required for structured output).

Example — save this as `rubric_v1.txt`:

```
You are an expert rater. Score the following text on two dimensions,
each on a 1–5 scale.

Dimension 1 — Clarity (1 = very unclear, 5 = very clear)
Dimension 2 — Emotional tone (1 = very negative, 5 = very positive)

Respond ONLY with a JSON object in this exact format and nothing else:
{"clarity": <1-5>, "emotional_tone": <1-5>}

Text to rate:
{{text}}
```

**The rubric is your scientific instrument.** Once you have collected data
against a version, never edit it in place. Save revisions under a new name
(`rubric_v2.txt`) so your output files always trace back to the exact rubric
that produced them.

## 4. Inspect the compiled prompt (no API cost)

Before spending any API credits, verify that the prompt assembles correctly:

```{r}
template_path    <- "rubric_v1.txt"
participant_text <- "The instructions were easy to follow and I felt great."

preview_prompt(read_template(template_path), participant_text)
```

This prints the full prompt with character and word counts. Read it as the
model will see it — confirm the participant text landed in the right place and
the rubric instructions are complete.

## 5. Score a single text interactively

Once the prompt looks right, test the full round-trip with one text:

```{r}
params <- run_params(model = "gpt-4o", temperature = 0)

result <- score_one(template_path, participant_text, params)
print_result(result$raw, result$scores)
```

`print_result()` shows both the raw JSON the model returned and the parsed
scores side by side. Ask yourself:

- Are all expected dimensions present?
- Are the values in the right range?
- Does the rating make intuitive sense for this text?

Repeat with two or three contrasting texts before moving on. This sanity check
costs a few API calls but can save you from discovering a rubric problem after
scoring 500 participants.

## 6. Prepare your data

`score_many()` expects a data frame with at minimum two columns:

| Column | Description |
|--------|-------------|
| `id`   | A unique identifier for each row (e.g. a participant ID) |
| `text` | The text to score |

```{r}
df <- read.csv("participants.csv", stringsAsFactors = FALSE)
head(df[, c("id", "text")])
```

The `id` column is what the skip logic uses to avoid rescoring rows that have
already been processed.

## 7. Scan and redact identifying information

Before sending any text to an external API, scan for potentially identifying
information (PII). This is especially important when participants provided
free-text responses.

```{r}
pii_result <- scan_pii(df)
```

`scan_pii()` separates findings into two tiers and prints a ready-to-use
action plan:

```
PII scan: 4 match(es) across 3 text(s)

Auto-redactable — use redact_pii() to clear:
  [P089] email
    "contact me at user@example.com"

Needs review — use redact_words() per row:
  [P023] name_disclosure #1
    "My name is Sarah Johnson and I…"
  [P023] name_disclosure #2
    "my friend's name is David and…"
  [P041] location_phrase #1
    "I live in Tel Aviv near the old…"

Suggested redact_words() calls — fill in n_words:
  df <- redact_words(df, id = "P023", pattern = "name_disclosure", n_words = ?, occurrence = 1)
  df <- redact_words(df, id = "P023", pattern = "name_disclosure", n_words = ?, occurrence = 2)
  df <- redact_words(df, id = "P041", pattern = "location_phrase", n_words = ?)
```

### Step 1 — Auto-redact structural PII

**Never edit your source data file.** Raw data is a scientific record; all
changes must live in code. `redact_pii()` replaces structural PII (emails,
phones, Prolific IDs, etc.) globally using the detection regex — the original
values never appear in your script:

```{r}
df <- redact_pii(df)
```

### Step 2 — Redact disclosure phrases per row

Disclosure patterns ("my name is", "I live in") capture only the trigger
phrase, not the name or place that follows. How many words to remove is a
judgment call per row. Copy the suggested calls from the `scan_pii()` output
and fill in `n_words`:

```{r}
df <- redact_words(df, id = "P023", pattern = "name_disclosure", n_words = 2, occurrence = 1)
df <- redact_words(df, id = "P023", pattern = "name_disclosure", n_words = 1, occurrence = 2)
df <- redact_words(df, id = "P041", pattern = "location_phrase",  n_words = 3)
```

`redact_words()` replaces the trigger plus `n_words` following words with
`[name disclosure redacted]`. When `occurrence` is omitted, all matches in
that row are replaced with the same word count.

### Skipping the check on subsequent runs

`score_many()` runs `scan_pii()` automatically (default `pii_check = TRUE`).
Once redaction is complete, set `pii_check = FALSE`:

```{r}
score_many(df, template_path, params,
           pii_check = FALSE, output_dir = output_dir)
```

### Customising the patterns

Add domain-specific patterns or remove noisy ones by modifying the two
pattern sets independently:

```{r}
auto <- auto_pii_patterns()
auto["national_id"] <- "\\b\\d{9}\\b"  # add Israeli national ID
auto <- auto[names(auto) != "url"]      # URLs not sensitive in this study

scan_pii(df, auto_patterns = auto)
redact_pii(df, patterns = auto)
```

## 8. Choose an output directory

`score_many()` and `status()` both take an `output_dir` argument. Results are
written to `<output_dir>/<rubric-stem>.csv` — so the filename always encodes
which rubric version produced it. Choose a directory that makes sense for your
project:

```{r}
output_dir <- "results"   # or "data", "output", an absolute path, etc.
```

The directory is created automatically if it does not exist.

## 9. Pilot run: score a small batch first

Always run a pilot before committing the full dataset. Use `n` to limit how
many rows are processed:

```{r}
score_many(df, template_path, params,
           n = 10, pii_check = FALSE, output_dir = output_dir)
```

Open `results/rubric_v1.csv` and check:

- Are all expected score columns present?
- Do the scores look reasonable across varied texts?
- Is there any evidence the rubric is being misread?

If something is wrong, fix the rubric (save it as `rubric_v2.txt`) and pilot
again before touching the full dataset.

## 10. Full batch run

Once the pilot looks good, run the full dataset. Rows already scored in the
pilot are automatically skipped:

```{r}
score_many(df, template_path, params,
           pii_check = FALSE, output_dir = output_dir)
```

A progress bar shows completion in real time. Individual failures are caught
per-row and written to `results/rubric_v1_errors.csv` — the run continues
rather than stopping. A summary is printed at the end:

```
Scoring 490 text(s) → results/rubric_v1.csv
[==========================================] 490/490 (100%)
Done. 488 scored, 2 failed — see results/rubric_v1_errors.csv.
```

Check status at any point:

```{r}
status(df, template_path, params, output_dir = output_dir)
# Rubric     : rubric_v1.txt
# Model      : gpt-4o
# Output dir : results
# Total      : 500
# Scored     : 488
# Failed     :   2
# Pending    :  10
```

## 11. Handling interruptions

If the run is interrupted (network error, crash, power loss), just run the
same command again. Completed rows are skipped, failed rows are retried:

```{r}
# Safe to run again at any time
score_many(df, template_path, params, output_dir = output_dir)
```

To process only the next batch of N rows:

```{r}
score_many(df, template_path, params, n = 50, output_dir = output_dir)
```

## 12. Large datasets: the OpenAI Batch API

For datasets of several hundred texts or more, the
[OpenAI Batch API](https://platform.openai.com/docs/guides/batch) is the
recommended approach. It submits all requests as a single asynchronous job
and returns results within 24 hours at **50% of the standard per-token cost**.

`score_many()` uses the synchronous API — one request at a time — which is
fine for pilots and moderate datasets but becomes slow and expensive at scale.
Batch API support (`submit_batch()` / `collect_batch()`) is planned for a
future version of this package.

## 13. Reading and using your results

Each row of the output CSV contains:

| Column | Description |
|--------|-------------|
| `id` | The participant identifier from your input data |
| `text` | The original text (useful for spot-checking) |
| *(score columns)* | One column per JSON key the model returned |
| `raw` | The exact JSON string from the model (for auditing) |
| `prompt_version` | Rubric filename stem (e.g. `rubric_v1`) |
| `model` | The model name (e.g. `gpt-4o`) |
| `temperature` | The temperature setting used |
| `scored_at` | ISO 8601 timestamp |

```{r}
results <- read.csv("results/rubric_v1.csv", stringsAsFactors = FALSE)

# Downstream analysis: aggregate scores however your study requires
results$overall <- rowMeans(results[, c("clarity", "emotional_tone")])
hist(results$overall)
```

The provenance columns (`prompt_version`, `model`, `scored_at`) make every row
self-describing — a CSV can be archived or shared and its origin is always
recoverable.
